A Linear Solution of the Linearity Challenge

### 2025-12-26 paper/calculus.tex
shortened poly echo server

\newcommand\MonoEchoS{\mathit{Server}}
\newcommand\MonoEchoC{\mathit{Client}}
\begin{example}
    \label{ex:neg}
    %
    We illustrate an example of higher-order session by modeling an echo server
    that consumes boolean values and sends them back unchanged. The protocol
    implemented by the server is described by the type $\OfCourse(\dual\Bool
    \Par (\Bool \Ten \One))$ where the modality~$\OfCourse$ indicates that the
    server is able to accept an unbounded number of requests and the type
    $\dual\Bool \Par (\Bool \Ten \One)$ describes the sequence of actions
    performed by the server at each connection with a client: the server first
    consumes a boolean value (say $u$, of type $\dual\Bool$), then produces
    another boolean value (say $v$, of type $\Bool$) and finally sends an empty
    message. We can model the server and a possible client thus:
    \begin{align*}
        \MonoEchoS(x) & \eqdef
        \Server\x\y.
        \Join\y\u{y'}.
        \Fork{y'}\v{y''}{
            \Link\v\u
        }{
            \Close{y''}
        } \\
        \MonoEchoC(x,z) & \eqdef
        \Client\x\y.
        \Fork\y\u{y'}{
            \TrueP(u)
        }{
            \Join{y'}\v{y''}.
            \Wait{y''}.
            \Link\v\z
        }
    \end{align*}

    Notice the use of continuations for chaining communications together. In the
    server, the link $\Link\v\u$ merges $v$ and $u$ so as to send the same
    channel $u$ received from the client. In the client, the link $\Link\v\z$
    ``assigns'' the message $v$ received from the server to the free channel
    $z$, which represents the result of the interaction.
    %
    If we write $\reds$ for the reflexive, transitive closure of $\red$ it is
    easy to verify that $\Cut\x{\MonoEchoC(x)}{\MonoEchoS(x)} \red^* \TrueP(z)$,
    as expected.
    %
    \eoe
\end{example}

@@

We use typing contexts (\ie sequents) to keep track of the type of channels.
Typing contexts are finite maps from channels to types written as $x_1 : A_1,
\dots, x_n : A_n$ and ranged over by $\ContextC$ and $\ContextD$. We write
$\ContextC, \ContextD$ for the union of $\ContextC$ and $\ContextD$ when they
have disjoint domains. We write $\WhyNot\Context$ for some context $\Context =
x_1 : \WhyNot A_1, \dots, x_n : \WhyNot A_n$ where all the types in its range
are prefixed by the $\WhyNot$ modality. We call these types \emph{unrestricted},
since they are used to denote shared channels that can be (explicitly) discarded
and duplicated at will.

\begin{table}[t]
    \caption{Typing rules of \Calculus.}
    \label{tab:typing-rules}
    \begin{tabular}{@{}c@{}}
        \begin{mathpar}
            \inferrule[\LinkRule]{~}{
                \wtp{\Link\x\y}{x : A, y : \dual{A}}
            }
            \and
            \inferrule[\FailRule]{~}{
                \wtp{\Fail\x}{x : \Top, \Context}
            }
            \and
            \inferrule[\WaitRule]{
                \wtp{P}{\Context}
            }{
                \wtp{\Wait\x.P}{x : \Bot, \Context}
            }
            \and
            \inferrule[\CloseRule]{~}{
                \wtp{\Close\x}{x : \One}
            }
            \and
            \inferrule[\CaseRule]{
                \wtp{P}{y : A, \Context}
                \\
                \wtp{Q}{y : B, \Context}
            }{
                \wtp{\Case\x\y{P}{Q}}{x : A \With B, \Context}
            }
            \and
            \inferrule[\SelectRule]{
                \wtp{P}{y : A_i, \Context}
            }{
                \wtp{\Select\x{\InTag_i}\y.P}{x : A_1 \Plus A_2, \Context}
            }
            \and
            \inferrule[\JoinRule]{
                \wtp{P}{y : A, z : B, \Context}
            }{
                \wtp{\Join\x\y\z.P}{x : A \Par B, \Context}
            }
            \and
            \inferrule[\ForkRule]{
                \wtp{P}{y : A, \ContextC}
                \\
                \wtp{Q}{z : B, \ContextD}
            }{
                \wtp{\Fork\x\y\z{P}{Q}}{x : A \Ten B, \ContextC, \ContextD}
            }
            \and
            \inferrule[\ExistsRule]{
                \wtp{P}{y : B\subst{A}\X, \Context}
            }{
                \wtp{\Exists\x{A}\y.P}{x : \exists\X.B, \Context}
            }
            \and
            \inferrule[\ForallRule]{
                \wtp{P}{y : A, \Context}
            }{
                \wtp{\Forall\x\X\y.P}{x : \forall\X.A, \Context}
            }
            ~X\not\in\fn\Context
            \and
            \inferrule[\ServerRule]{
                \wtp{P}{y : A, \WhyNot\Context}
            }{
                \wtp{\Server\x\y.P}{x : \OfCourse A, \WhyNot\Context}
            }
            \and
            \inferrule[\ClientRule]{
                \wtp{P}{y : A, \Context}
            }{
                \wtp{\Client\x\y.P}{x : \WhyNot A, \Context}
            }
            \and
            \inferrule[\WeakenRule]{
                \wtp{P}\Context
            }{
                \wtp{P}{x : \WhyNot A, \Context}
            }
            \and
            \inferrule[\ContractRule]{
                \wtp{P}{y : \WhyNot A, z : \WhyNot A, \Context}
            }{
                \wtp{P}{x : \WhyNot A, \Context}
            }
            \and
            \inferrule[\CutRule]{
                \wtp{P}{x : A, \ContextC}
                \\
                \wtp{Q}{x : \dual{A}, \ContextD}
            }{
                \wtp{\Cut[A]\x{P}{Q}}{\ContextC, \ContextD}
            }
        \end{mathpar}
    \end{tabular}
\end{table}

Typing judgments have the form $\wtp{P}\Context$ meaning that the process $P$ is
well typed in the context $\Context$. Equivalently, the judgment indicates that
the sequent $\wtp{}\Context$ is derivable and $P$ is a \emph{proof term}
corresponding to the derivation for $\wtp{}\Context$.
%
The typing rules are shown in \Cref{tab:typing-rules}. Since they are in
one-to-one correspondence with -- and have exactly the same structure of -- the
proof rules of classical linear logic, the reader may refer to the standard
literature on propositions as sessions~\citep{Wadler14} for their
interpretation.

@@

\begin{example}
    \label{ex:echo}
    In this example we define a polymorphic echo server:
    \[
        \EchoP(x) \eqdef
        \Server\x\y.
        \Forall\y\X{y'}.
        \Join{y'}\u{y''}.
        \Fork{y''}\v{y'''}{\Link\u\v}{\Close{y'''}}
    \]

    The server accepts requests on the shared channel $x$. Each request creates
    a new session, initially represented by the channel $y$. The session
    consists in receiving the type $X$ of the value being echoed followed by the
    value itself $u$. The value is then sent back unchanged and the session is
    terminated. Note the succession of different (linear) channels $y, y', y'',
    y'''$ that chain the session together.

    We establish that $\EchoP(x)$ is well typed with the following derivation
    \[
        \begin{prooftree}
            \[
                \[
                    \[
                        \[
                            \justifies
                            \wtp{
                                \Link\u\v
                            }{
                                u : X,
                                v : \dual\X
                            }
                            \using\LinkRule
                        \]
                        \[
                            \justifies
                            \wtp{
                                \Close{y'''}
                            }{
                                y''' : \One
                            }
                            \using\CloseRule
                        \]
                        \justifies
                        \wtp{
                            \Fork{y''}\v{y'''}{\Link\u\v}{\Close{y'''}}
                        }{
                            u : X,
                            y'' : \dual\X \Ten \One
                        }
                        \using\ForkRule
                    \]
                    \justifies
                    \wtp{
                        \Join{y'}\u{y''}.
                        \Fork{y''}\v{y'''}{\Link\u\v}{\Close{y'''}}
                    }{
                        y' : X \Par (\dual\X \Ten \One)
                    }
                    \using\JoinRule
                \]
                \justifies
                \wtp{
                    \Forall\y\X{y'}.
                    \Join{y'}\u{y''}.
                    \Fork{y''}\v{y'''}{\Link\u\v}{\Close{y'''}}
                }{
                    y : \forall\X.X \Par (\dual\X \Ten \One)
                }
                \using\ForallRule
            \]
            \justifies
            \wtp{
                \EchoP(x)
            }{
                x : \OfCourse(\forall\X.X \Par (\dual\X \Ten \One))
            }
            \using\ServerRule
        \end{prooftree}
    \]
    %
    where the side condition of the rule \ForallRule is trivially satifised
    since the typing context does not contain bindings other than the one for
    $y$.
    %
    \eoe
\end{example}

>>>>

\begin{example}
    \label{ex:neg}
    %
    We illustrate an example of higher-order session by modeling an echo server
    that consumes boolean values and sends them back unchanged. The protocol
    implemented by the server is described by the type $\OfCourse(\dual\Bool
    \Par (\Bool \Ten \One))$ where the modality~$\OfCourse$ indicates that the
    server is able to accept an unbounded number of requests and the type
    $\dual\Bool \Par (\Bool \Ten \One)$ describes the sequence of actions
    performed by the server at each connection with a client: the server first
    consumes a boolean value (say $u$, of type $\dual\Bool$), then produces
    another boolean value (say $v$, of type $\Bool$) and finally sends an empty
    message. We can model the server and a possible client thus:
    \begin{align*}
        \ServerP(x) & \eqdef
        \Server\x\y.
        \Join\y\u{y'}.
        \Fork{y'}\v{y''}{
            \Link\v\u
        }{
            \Close{y''}
        } \\
        \ClientP(x,z) & \eqdef
        \Client\x\y.
        \Fork\y\u{y'}{
            \TrueP(u)
        }{
            \Join{y'}\v{y''}.
            \Wait{y''}.
            \Link\v\z
        }
    \end{align*}

    Notice the use of continuations for chaining communications together. In the
    server, the link $\Link\v\u$ merges $v$ and $u$ so as to send the same
    channel $u$ received from the client. In the client, the link $\Link\v\z$
    ``assigns'' the message $v$ received from the server to the free channel
    $z$, which represents the result of the interaction.
    %
    If we write $\reds$ for the reflexive, transitive closure of $\red$ it is
    easy to verify that $\Cut\x{\ClientP(x)}{\ServerP(x)} \red^* \TrueP(z)$,
    as expected.
    %
    \eoe
\end{example}

@@

We use typing contexts (\ie sequents) to keep track of the type of channels.
Typing contexts are finite maps from channels to types written as $x_1 : A_1,
\dots, x_n : A_n$ and ranged over by $\ContextC$ and $\ContextD$. We write
$\ContextC, \ContextD$ for the union of $\ContextC$ and $\ContextD$ when they
have disjoint domains. We write $\WhyNot\Context$ for some context $\Context =
x_1 : \WhyNot A_1, \dots, x_n : \WhyNot A_n$ where all the types in its range
are prefixed by the modality~$\WhyNot$. We call these types \emph{unrestricted}
because they are used to denote shared channels that can be (explicitly)
discarded and duplicated.

\begin{table}[t]
    \caption{Typing rules of \Calculus.}
    \label{tab:typing-rules}
    \begin{tabular}{@{}c@{}}
        \begin{mathpar}
            \inferrule[\LinkRule]{~}{
                \wtp{\Link\x\y}{x : A, y : \dual{A}}
            }
            \and
            \inferrule[\FailRule]{~}{
                \wtp{\Fail\x}{x : \Top, \Context}
            }
            \and
            \inferrule[\WaitRule]{
                \wtp{P}{\Context}
            }{
                \wtp{\Wait\x.P}{x : \Bot, \Context}
            }
            \and
            \inferrule[\CloseRule]{~}{
                \wtp{\Close\x}{x : \One}
            }
            \and
            \inferrule[\CaseRule]{
                \wtp{P}{y : A, \Context}
                \\
                \wtp{Q}{y : B, \Context}
            }{
                \wtp{\Case\x\y{P}{Q}}{x : A \With B, \Context}
            }
            \and
            \inferrule[\SelectRule]{
                \wtp{P}{y : A_i, \Context}
            }{
                \wtp{\Select\x{\InTag_i}\y.P}{x : A_1 \Plus A_2, \Context}
            }
            \and
            \inferrule[\JoinRule]{
                \wtp{P}{y : A, z : B, \Context}
            }{
                \wtp{\Join\x\y\z.P}{x : A \Par B, \Context}
            }
            \and
            \inferrule[\ForkRule]{
                \wtp{P}{y : A, \ContextC}
                \\
                \wtp{Q}{z : B, \ContextD}
            }{
                \wtp{\Fork\x\y\z{P}{Q}}{x : A \Ten B, \ContextC, \ContextD}
            }
            \and
            \inferrule[\ExistsRule]{
                \wtp{P}{y : B\subst{A}\X, \Context}
            }{
                \wtp{\Exists\x{A}\y.P}{x : \exists\X.B, \Context}
            }
            \and
            \inferrule[\ForallRule]{
                \wtp{P}{y : A, \Context}
            }{
                \wtp{\Forall\x\X\y.P}{x : \forall\X.A, \Context}
            }
            ~X\not\in\fn\Context
            \and
            \inferrule[\ServerRule]{
                \wtp{P}{y : A, \WhyNot\Context}
            }{
                \wtp{\Server\x\y.P}{x : \OfCourse A, \WhyNot\Context}
            }
            \and
            \inferrule[\ClientRule]{
                \wtp{P}{y : A, \Context}
            }{
                \wtp{\Client\x\y.P}{x : \WhyNot A, \Context}
            }
            \and
            \inferrule[\WeakenRule]{
                \wtp{P}\Context
            }{
                \wtp{\Weaken\x.P}{x : \WhyNot A, \Context}
            }
            \and
            \inferrule[\ContractRule]{
                \wtp{P}{y : \WhyNot A, z : \WhyNot A, \Context}
            }{
                \wtp{\Contract\x\y\z.P}{x : \WhyNot A, \Context}
            }
            \and
            \inferrule[\CutRule]{
                \wtp{P}{x : A, \ContextC}
                \\
                \wtp{Q}{x : \dual{A}, \ContextD}
            }{
                \wtp{\Cut[A]\x{P}{Q}}{\ContextC, \ContextD}
            }
        \end{mathpar}
    \end{tabular}
\end{table}

Typing judgments have the form $\wtp{P}\Context$ meaning that the process $P$ is
well typed in the context $\Context$. Equivalently, the judgment indicates that
the sequent $\wtp{}\Context$ is derivable and $P$ is a proof term corresponding
to the derivation for $\wtp{}\Context$.
%
The typing rules are shown in \Cref{tab:typing-rules}. They are in one-to-one
correspondence with -- and have exactly the same structure of -- the proof rules
of classical linear logic.
%
The reader may refer to the standard literature on propositions as
sessions~\citep{Wadler14,LindleyMorris15,GayVasconcelos25} for the
interpretation of the rules. The only relevant difference with the typing rules
in these papers is that the premises mentioning the continuation channel $z$
actually refer to the same channel $x$ on which the process in the conclusion is
acting.

@@

\begin{example}
    \label{ex:echo}
    Looking at \Cref{ex:echo} we notice that the server does not make any
    assumption on the type of the values it receives and sends. Therefore, we
    can define a polymorphic version of the echo server that works for every
    message type and not just for the booleans.
    %
    The polymorphic version of the echo server is defined below:
    \[
        \ServerP(x) \eqdef
        \Server\x\y.
        \Forall\y\X{y'}.
        \Join{y'}\u{y''}.
        \Fork{y''}\v{y'''}{\Link\u\v}{\Close{y'''}}
    \]

    The only difference with respect to the server in \Cref{ex:echo} is that now
    the process receives the type $X$ of the messages to be processed and then
    continues as before.
    %
    We establish that $\ServerP(x)$ is well typed with the following derivation
    \[
        \begin{prooftree}
            \[
                \[
                    \[
                        \[
                            \justifies
                            \wtp{
                                \Link\u\v
                            }{
                                u : \dual\X,
                                v : X
                            }
                            \using\LinkRule
                        \]
                        \[
                            \justifies
                            \wtp{
                                \Close{y'''}
                            }{
                                y''' : \One
                            }
                            \using\CloseRule
                        \]
                        \justifies
                        \wtp{
                            \Fork{y''}\v{y'''}{\Link\u\v}{\Close{y'''}}
                        }{
                            u : \dual\X,
                            y'' : X \Ten \One
                        }
                        \using\ForkRule
                    \]
                    \justifies
                    \wtp{
                        \Join{y'}\u{y''}.
                        \Fork{y''}\v{y'''}{\Link\u\v}{\Close{y'''}}
                    }{
                        y' : \dual\X \Par (X \Ten \One)
                    }
                    \using\JoinRule
                \]
                \justifies
                \wtp{
                    \Forall\y\X{y'}.
                    \Join{y'}\u{y''}.
                    \Fork{y''}\v{y'''}{\Link\u\v}{\Close{y'''}}
                }{
                    y : \forall\X.\dual\X \Par (X \Ten \One)
                }
                \using\ForallRule
            \]
            \justifies
            \wtp{
                \ServerP(x)
            }{
                x : \OfCourse(\forall\X.\dual\X \Par (X \Ten \One))
            }
            \using\ServerRule
        \end{prooftree}
    \]
    %
    where the side condition of the rule \ForallRule is trivially satifised
    since the typing context does not contain bindings other than the one for
    $y$.
    %
    \eoe
\end{example}
